---
title: "data_visualization_hackathon"
author: "Sohrab Khan"
date: "2024-07-26"
output:
  pdf_document: default
  html_document:
    highlight: tango
    theme: cerulean
---
# **Hackathon: Data Visualization in R**

## **Load and Install Repositories and Packages**
```{r}
library(tidyverse)
library(ggplot2)
library(readxl)
library(dplyr)
```
## **Create a data folder manually or use R code mentioned below**
```{r}
dir.create("hackathon_data_visualization")
```
## **Download example data**
```{r}
url <- "https://raw.githubusercontent.com/AMMnet/AMMnet-Hackathon/main/01_data-vis/data/"
download.file(paste0(url,"mockdata_cases.csv"),
              destfile = "hackathon_data_visualization/mockdata_cases.csv")
download.file("https://raw.githubusercontent.com/AMMnet/AMMnet-Hackathon/main/01_data-vis/data/mosq_mock.csv", destfile = "hackathon_data_visualization/mosq_mock.csv")
```
## **Load example data**
```{r}
library(readr)
malaria_data <- read_csv("hackathon_data_visualization/mockdata_cases.csv")
View(malaria_data)
mosq_data <- read_csv("hackathon_data_visualization/mosq_mock.csv")
View(mosq_data)
```
## **Characterize means Exploring our data**
### **How many rows and columns**
```{r}
dim(malaria_data)
```

### **General information**
```{r}
head(malaria_data)
summary(malaria_data)
```
### **Exploring individual columns of the data**
```{r}
malaria_data$location # values for a single column
unique(malaria_data$location) # unique values for a single column
table(malaria_data$location) # frequencies for a single column
table(malaria_data$location, malaria_data$ages) # frequencies for multiple column
```
### **Check for if there are any missing values- NA**
```{r}
sum(is.na(malaria_data))
```
## **Exploratory Visualizations Using Base R Functions**
### **Single variable or column comparison**
#### **Histogram or Frequency Chart
```{r}
hist(malaria_data$prev)
hist(malaria_data$prev,
     breaks = 10, # breaks mean how many individual bars do we need to group
     main = "Distribution of Malaria Prevalence",
     xlab = "Malaria Prevalence",
     ylab = "Frequency",
     col = "#701f28",
     border = "black")
```

#### **Barplot- tell us the number of counts within a categorical variable/ column**
```{r}
barplot(table(malaria_data$location))
barplot(table(malaria_data$year))
```

#### **Plotting Multiple column/ variable**
##### **Scatterplot using R code - plot**
```{r}
plot(x = malaria_data$total, y = malaria_data$positive)
plot(x = malaria_data$month, y = malaria_data$prev)
```

##### **Boxplot using R code - boxplot**
```{r}
boxplot(malaria_data$prev ~ malaria_data$location)
boxplot(malaria_data$prev ~ malaria_data$month)
boxplot(malaria_data$prev ~ malaria_data$month,
        data = malaria_data,
        xlab = "Malaria Month",
        ylab = "Malaria Prevalence",
        col = "#701f28",
        border = "black")
jan_data <- filter(malaria_data, month==1)
boxplot(jan_data$prev ~ jan_data$month)
boxplot
```
## **Step by Step Data Visualization using ggplot2**
### **Data**
```{r}
ggplot(data = malaria_data)
```

### **Organize Data(aes())**
```{r}
ggplot(data = malaria_data, aes(x = total, y = positive))
```

### **Visualizing Data(geom_)**
```{r}
ggplot(data = malaria_data, aes(x = total, y = positive)) + geom_point()
ggplot(data = malaria_data, aes(x = positive)) + geom_histogram(bins = 25)
ggplot(data = malaria_data, aes(x = month)) + geom_bar(fill = "tomato")
```

### **Add multiple geoms_**
```{r}
ggplot(data = malaria_data, aes(x = location, y = prev)) + geom_boxplot()+
  geom_jitter(alpha = 0.2) #Boxplot shows value for the range but they don't show distribution
ggplot(data = malaria_data, aes(x = location, y = prev)) + geom_violin(fill= "darkorchid2")+
  geom_jitter(alpha = 0.2) #Similar to the boxplot but the shape is different shows distribution
ggplot(data = malaria_data, aes(x = total, y = positive)) + geom_point() +  geom_smooth(method = "lm")
```

### **Extending the aesthetic**
```{r}
ggplot(data = malaria_data, aes(x = total, y = positive, colour = location)) + geom_point()
ggplot(data = malaria_data, aes(x = prev, fill = ages)) + geom_histogram(colour = "black", bins = 17)
ggplot(data = malaria_data, aes(x = location, y = prev, fill = location)) + geom_boxplot()+
  geom_jitter(alpha = 0.2)
ggplot(data = malaria_data, aes(x = total, y = positive, colour = location)) + geom_point()+ geom_smooth(method = "lm", se= FALSE) + theme_classic()
ggplot(data = malaria_data, aes(x = location, y = prev, fill = location)) + geom_boxplot()+
  geom_jitter(alpha = 0.2) +theme_classic() + theme(legend.position = "bottom")
```

#### **Add colour Palette automatic using scale fill brewer**
```{r}
ggplot(data = malaria_data, aes(x = location, y = prev, fill = location)) + geom_boxplot()+
  geom_jitter(alpha = 0.2) +theme_classic() + scale_fill_brewer(palette = "RdPu")
```

#### **Add colour manually**
```{r}
ggplot(data = malaria_data, aes(x = location, y = prev, fill = location)) + geom_boxplot()+
  geom_jitter(alpha = 0.2) +theme_classic() + scale_fill_manual(values = c("chartreuse", "chartreuse1", "chartreuse2","chartreuse3", "chartreuse4"))
```

#### **Use viridis package to create custom color palettes**
```{r}
library(viridis)
ggplot(data = malaria_data, aes(x = total, y = positive, colour = prev)) + geom_point() +
  scale_color_viridis(option = "magma")
```

#### **Small multiple plots using facet**
```{r}
ggplot(data = malaria_data, aes(x = total, y = positive, colour = prev)) + geom_point() +
  scale_color_viridis(option = "magma") +
  facet_wrap(~location)+theme_classic()
```

#### **Visualize continous data on a spectrum using scale color gradient**
```{r}
ggplot(data = malaria_data, aes(x = total, y = positive, colour = prev)) + geom_point() +
  scale_color_gradient(low = "red", high = "#701f28")
```

#### **Adding title, subtitle and flipping x and y axis**
```{r}
ggplot(data = malaria_data, aes(x = location, y = prev, fill = location)) + geom_boxplot()+
  geom_jitter(alpha = 0.2) +
  facet_wrap(~ages) +
  coord_flip() + # flips the x and y axis
  theme_classic() + scale_fill_brewer(palette = "RdPu")+
  labs(title = "Malaria Prevalence by Location and Age",
       subtitle = "Data from 2018 - 2020",
       x = "Location",
       y = "Prevalence",
       fill = "Age group")
ggsave("malaria_age_prevalenceboxplot.png", width = 10, height = 6, units = "in", dpi = 300)
```
